## Conserved Taxonomy

*  Input file: *.nr.wang.pick.pick.tx.1.cons.tax.summary* file. 
* All chimeras, non-bacterial lineages and mock sequences have been removed

```{r constaxsummary, echo=TRUE, message=FALSE, warning=FALSE}
taxsummary = read.table(file="../data/mothur/phylotype/qced.trim.contigs.good.unique.good.filter.unique.precluster.pick.nr.wang.pick.pick.tx.1.cons.tax.summary", stringsAsFactors=FALSE, header=T, sep='\t')

# Save taxsummary object 
write.table(taxsummary, "../results/tables/phylotype/conserved_taxonomy/taxsummary.tsv", append = FALSE, quote = FALSE, sep="\t", row.names = FALSE, col.names = TRUE)
```

**Format of *.nr.wang.pick.pick.tx.1.cons.tax.summary* file**
```{r}
library(kableExtra)

knitr::kable(head(taxsummary[,1:10], 10), "html"
) %>%
  kable_styling(bootstrap_options = "striped", font_size = 10)
```


### Taxon dataframes
* Files with extension **.a** means **all data** incluidng classified and unclassified
```{r, echo=TRUE}
library(dplyr)

phylum.a <- taxsummary %>% filter(grepl("2", taxlevel))
phylum.a <- phylum.a[, -c(1,2,4,5)]

class.a <- taxsummary %>% filter(grepl("3", taxlevel))
class.a <- class.a[, -c(1,2,4,5)]

order.a <- taxsummary %>% filter(grepl("4", taxlevel))
order.a <- order.a[, -c(1,2,4,5)]

family.a <- taxsummary %>% filter(grepl("5", taxlevel))
family.a <- family.a[, -c(1,2,4,5)]

genus.a <- taxsummary %>% filter(grepl("6", taxlevel))
genus.a <- genus.a[, -c(1,2,4,5)]
```

**Format of *phylum.a* file**
```{r}
library(kableExtra)

knitr::kable(head(phylum.a[,1:10], 10), "html"
) %>%
  kable_styling(bootstrap_options = "striped", font_size = 10)
```


### Dimension of unfiltered taxa objects
```{r, echo=TRUE}
dim(phylum.a)
dim(class.a)
dim(order.a)
dim(family.a)
dim(genus.a)
```

### Save each taxon dataframe to a file
```{r, echo=TRUE}
write.table(phylum.a,"../results/tables/phylotype/conserved_taxonomy/phylum.a.tsv", quote = FALSE, sep="\t",
            row.names = FALSE, col.names = TRUE)

write.table(class.a,"../results/tables/phylotype/conserved_taxonomy/class.a.tsv", quote = FALSE, sep="\t",
            row.names = FALSE, col.names = TRUE)

write.table(order.a,"../results/tables/phylotype/conserved_taxonomy/order.a.tsv", quote = FALSE, sep="\t",
            row.names = FALSE, col.names = TRUE)

write.table(family.a,"../results/tables/phylotype/conserved_taxonomy/family.a.tsv", quote = FALSE, sep="\t",
            row.names = FALSE, col.names = TRUE)

write.table(genus.a,"../results/tables/phylotype/conserved_taxonomy/genus.a.tsv", quote = FALSE, sep="\t",
            row.names = FALSE, col.names = TRUE)
```

<br>

### Remove unclassified taxa
```{r, echo=TRUE}
library(dplyr)
phylum <- phylum.a %>% filter(!grepl("*unclassified", taxon))
class <- class.a %>% filter(!grepl("*unclassified", taxon))
order <- order.a %>% filter(!grepl("*unclassified", taxon))
family <- family.a %>% filter(!grepl("*unclassified", taxon))
genus <- genus.a %>% filter(!grepl("*unclassified", taxon))

dim(phylum)
dim(class)
dim(order)
dim(family)
dim(genus)
```

### Sort taxon in ascending order
```{r, echo=TRUE}
phylum<- phylum[order(phylum$taxon),]
class<- class[order(class$taxon),]
order<- order[order(order$taxon),]
family<- family[order(family$taxon),]
genus<- genus[order(genus$taxon),]
```


*Sorted Phylum*
```{r}
library(kableExtra)

knitr::kable(head(phylum[,1:10], 10), "html"
) %>%
  kable_styling(bootstrap_options = "striped", font_size = 10)
```

### Save sorted taxa objects to files
```{r, echo=TRUE}
write.table(phylum,"../results/tables/phylotype/conserved_taxonomy/phylum.tsv", quote = FALSE, sep="\t",
            row.names = FALSE, col.names = TRUE)

write.table(class,"../results/tables/phylotype/conserved_taxonomy/class.tsv", quote = FALSE, sep="\t",
            row.names = FALSE, col.names = TRUE)

write.table(order,"../results/tables/phylotype/conserved_taxonomy/order.tsv", quote = FALSE, sep="\t",
            row.names = FALSE, col.names = TRUE)

write.table(family,"../results/tables/phylotype/conserved_taxonomy/family.tsv", quote = FALSE, sep="\t",
            row.names = FALSE, col.names = TRUE)

write.table(genus,"../results/tables/phylotype/conserved_taxonomy/genus.tsv", quote = FALSE, sep="\t",
            row.names = FALSE, col.names = TRUE)
```

### Transpose: Column to row
* Note that all samples are organized by column and the taxon by row.
* Most functions need samples organized by row so we will transpose the table
* Optionally to transpose correclty we will strip off the sample identifiers and convert them to row names so that the data matrix contains only numeric count data. This can vary greatly:)
* We will do this using *setNames* function and indicate [,-1] which means exclude column 1 currently containing the taxon.

```{r, echo=TRUE}
phylum<-read.table(file="../results/tables/phylotype/conserved_taxonomy/phylum.tsv", header = TRUE, sep="\t")
phylum.t = setNames(data.frame(t(phylum[,-1])), phylum[,1])
class<-read.table(file="../results/tables/phylotype/conserved_taxonomy/class.tsv", header = TRUE, sep="\t")
class.t = setNames(data.frame(t(class[,-1])), class[,1])
order<-read.table(file="../results/tables/phylotype/conserved_taxonomy/order.tsv", header = TRUE, sep="\t")
order.t = setNames(data.frame(t(order[,-1])), order[,1])
family<-read.table(file="../results/tables/phylotype/conserved_taxonomy/family.tsv", header = TRUE, sep="\t")
family.t = setNames(data.frame(t(family[,-1])), family[,1])
genus<-read.table(file="../results/tables/phylotype/conserved_taxonomy/genus.tsv", header = TRUE, sep="\t")
genus.t = setNames(data.frame(t(genus[,-1])), genus[,1])
```

<br>

```{r, echo=TRUE, comment=NA}
dim(phylum.t)
dim(class.t)
dim(order.t)
dim(family.t)
dim(genus.t)
```

**Format of transposed data**
```{r, echo=TRUE}
library(kableExtra)

knitr::kable(head(phylum.t[,1:10], 10), "html"
) %>%
  kable_styling(bootstrap_options = "striped", font_size = 10)
```

<br>

### Save transposed data
```{r, echo=TRUE}
write.table(phylum.t,"../results/tables/phylotype/conserved_taxonomy/phylum.t.tsv", append = FALSE, quote = FALSE, sep="\t", eol = "\n", na = "NA", dec = ".", row.names = FALSE, col.names = TRUE, qmethod = c("escape", "double"))
write.table(class.t,"../results/tables/phylotype/conserved_taxonomy/class.t.tsv", append = FALSE, quote = FALSE, sep="\t", eol = "\n", na = "NA", dec = ".", row.names = FALSE, col.names = TRUE, qmethod = c("escape", "double"))
write.table(order.t,"../results/tables/phylotype/conserved_taxonomy/order.t.tsv", append = FALSE, quote = FALSE, sep="\t", eol = "\n", na = "NA", dec = ".", row.names = FALSE, col.names = TRUE, qmethod = c("escape", "double"))
write.table(family.t,"../results/tables/phylotype/conserved_taxonomy/family.t.tsv", append = FALSE, quote = FALSE, sep="\t", eol = "\n", na = "NA", dec = ".", row.names = FALSE, col.names = TRUE, qmethod = c("escape", "double"))
write.table(genus.t,"../results/tables/phylotype/conserved_taxonomy/genus.t.tsv", append = FALSE, quote = FALSE, sep="\t", eol = "\n", na = "NA", dec = ".", row.names = FALSE, col.names = TRUE, qmethod = c("escape", "double"))
```

<br>

### Get relative abundance proportions
```{r, echo=TRUE}
phylumprop <- phylum.t/rowSums(phylum.t)
classprop <- class.t/rowSums(class.t)
orderprop <- order.t/rowSums(order.t)
familyprop <- family.t/rowSums(family.t)
genusprop <- genus.t/rowSums(genus.t)
```

<br>

```{r, echo=TRUE, comment=NA}
dim(phylumprop)
dim(classprop)
dim(orderprop)
dim(familyprop)
dim(genusprop)
```

**Format of relative phylum**
```{r, echo=TRUE}
library(kableExtra)

knitr::kable(head(phylumprop[,1:10], 10), "html"
) %>%
  kable_styling(bootstrap_options = "striped", font_size = 10)
```
<br>

### Save relative abundance objects
```{r, echo=TRUE}
write.table(phylumprop,"../results/tables/phylotype/conserved_taxonomy/phylumprop.tsv", append = FALSE, quote = FALSE, sep="\t", eol = "\n", na = "NA", dec = ".", row.names = FALSE, col.names = TRUE, qmethod = c("escape", "double"))
write.table(classprop,"../results/tables/phylotype/conserved_taxonomy/classprop.tsv", append = FALSE, quote = FALSE, sep="\t", eol = "\n", na = "NA", dec = ".", row.names = FALSE, col.names = TRUE, qmethod = c("escape", "double"))
write.table(orderprop,"../results/tables/phylotype/conserved_taxonomy/orderprop.tsv", append = FALSE, quote = FALSE, sep="\t", eol = "\n", na = "NA", dec = ".", row.names = FALSE, col.names = TRUE, qmethod = c("escape", "double"))
write.table(familyprop,"../results/tables/phylotype/conserved_taxonomy/familyprop.tsv", append = FALSE, quote = FALSE, sep="\t", eol = "\n", na = "NA", dec = ".", row.names = FALSE, col.names = TRUE, qmethod = c("escape", "double"))
write.table(genusprop,"../results/tables/phylotype/conserved_taxonomy/genusprop.tsv", append = FALSE, quote = FALSE, sep="\t", eol = "\n", na = "NA", dec = ".", row.names = FALSE, col.names = TRUE, qmethod = c("escape", "double"))
```

<br>

### Review all dimensions of taxa data
```{r, echo=FALSE, comment=NA}
message("Inludes unclassified")
dim(phylum.a)
dim(class.a)
dim(order.a)
dim(family.a)
dim(genus.a)
message("\n")

message("Only classified")
dim(phylum)
dim(class)
dim(order)
dim(family)
dim(genus)
message("\n")

message("Transposed")
dim(phylum.t)
dim(class.t)
dim(order.t)
dim(family.t)
dim(genus.t)
message("\n")

message("Relative proportions")
dim(phylumprop)
dim(classprop)
dim(orderprop)
dim(familyprop)
dim(genusprop)
```

### Create taxon objects containing variables

```{r, echo=TRUE}
phylum.t.v <- phylum.t
phylum.t.v$SampleCode <- as.character(bushmeatvars$SampleCode)
phylum.t.v$Condition <- as.character(bushmeatvars$Condition)
phylum.t.v$Species <- as.character(bushmeatvars$Species)
phylum.t.v$Season <- as.character(bushmeatvars$Season)
phylum.t.v$Ecosystem <- as.character(bushmeatvars$Ecosystem)

phylumprop.v <- phylumprop
phylumprop.v$SampleCode <- as.character(bushmeatvars$SampleCode)
phylumprop.v$Condition <- as.character(bushmeatvars$Condition)
phylumprop.v$Species <- as.character(bushmeatvars$Species)
phylumprop.v$Season <- as.character(bushmeatvars$Season)
phylumprop.v$Ecosystem <- as.character(bushmeatvars$Ecosystem)

class.t.v <- class.t
class.t.v$SampleCode <- as.character(bushmeatvars$SampleCode)
class.t.v$Condition <- as.character(bushmeatvars$Condition)
class.t.v$Species <- as.character(bushmeatvars$Species)
class.t.v$Season <- as.character(bushmeatvars$Season)
class.t.v$Ecosystem <- as.character(bushmeatvars$Ecosystem)

classprop.v <- classprop
classprop.v$SampleCode <- as.character(bushmeatvars$SampleCode)
classprop.v$Condition <- as.character(bushmeatvars$Condition)
classprop.v$Species <- as.character(bushmeatvars$Species)
classprop.v$Season <- as.character(bushmeatvars$Season)
classprop.v$Ecosystem <- as.character(bushmeatvars$Ecosystem)

order.t.v <- order.t
order.t.v$SampleCode <- as.character(bushmeatvars$SampleCode)
order.t.v$Condition <- as.character(bushmeatvars$Condition)
order.t.v$Species <- as.character(bushmeatvars$Species)
order.t.v$Season <- as.character(bushmeatvars$Season)
order.t.v$Ecosystem <- as.character(bushmeatvars$Ecosystem)

orderprop.v <- orderprop
orderprop.v$SampleCode <- as.character(bushmeatvars$SampleCode)
orderprop.v$Condition <- as.character(bushmeatvars$Condition)
orderprop.v$Species <- as.character(bushmeatvars$Species)
orderprop.v$Season <- as.character(bushmeatvars$Season)
orderprop.v$Ecosystem <- as.character(bushmeatvars$Ecosystem)

family.t.v <- family.t
family.t.v$SampleCode <- as.character(bushmeatvars$SampleCode)
family.t.v$Condition <- as.character(bushmeatvars$Condition)
family.t.v$Species <- as.character(bushmeatvars$Species)
family.t.v$Season <- as.character(bushmeatvars$Season)
family.t.v$Ecosystem <- as.character(bushmeatvars$Ecosystem)

familyprop.v <- familyprop
familyprop.v$SampleCode <- as.character(bushmeatvars$SampleCode)
familyprop.v$Condition <- as.character(bushmeatvars$Condition)
familyprop.v$Species <- as.character(bushmeatvars$Species)
familyprop.v$Season <- as.character(bushmeatvars$Season)
familyprop.v$Ecosystem <- as.character(bushmeatvars$Ecosystem)

genus.t.v <- genus.t
genus.t.v$SampleCode <- as.character(bushmeatvars$SampleCode)
genus.t.v$Condition <- as.character(bushmeatvars$Condition)
genus.t.v$Species <- as.character(bushmeatvars$Species)
genus.t.v$Season <- as.character(bushmeatvars$Season)
genus.t.v$Ecosystem <- as.character(bushmeatvars$Ecosystem)

genusprop.v <- genusprop
genusprop.v$SampleCode <- as.character(bushmeatvars$SampleCode)
genusprop.v$Condition <- as.character(bushmeatvars$Condition)
genusprop.v$Species <- as.character(bushmeatvars$Species)
genusprop.v$Season <- as.character(bushmeatvars$Season)
genusprop.v$Ecosystem <- as.character(bushmeatvars$Ecosystem)
```

### Save taxa objects containing variables
* Adding metadata (variables) to the objects makes facilitates simplifies beta diversity analysis.
```{r, echo=TRUE}
write.table(phylum.t.v,"../results/tables/phylotype/conserved_taxonomy/phylum.t.v.tsv", quote = FALSE, sep="\t",
            row.names = FALSE, col.names = TRUE)

write.table(class.t.v,"../results/tables/phylotype/conserved_taxonomy/class.t.v.tsv", quote = FALSE, sep="\t",
            row.names = FALSE, col.names = TRUE)

write.table(order.t.v,"../results/tables/phylotype/conserved_taxonomy/order.t.v.tsv", quote = FALSE, sep="\t",
            row.names = FALSE, col.names = TRUE)

write.table(family.t.v,"../results/tables/phylotype/conserved_taxonomy/family.t.v.tsv", quote = FALSE, sep="\t",
            row.names = FALSE, col.names = TRUE)

write.table(genus.t.v,"../results/tables/phylotype/conserved_taxonomy/genus.t.v.tsv", quote = FALSE, sep="\t",
            row.names = FALSE, col.names = TRUE)



write.table(phylumprop.v,"../results/tables/phylotype/conserved_taxonomy/phylumprop.v.tsv", quote = FALSE, sep="\t",
            row.names = FALSE, col.names = TRUE)

write.table(classprop.v,"../results/tables/phylotype/conserved_taxonomy/classprop.v.tsv", quote = FALSE, sep="\t",
            row.names = FALSE, col.names = TRUE)

write.table(orderprop.v,"../results/tables/phylotype/conserved_taxonomy/orderprop.v.tsv", quote = FALSE, sep="\t",
            row.names = FALSE, col.names = TRUE)

write.table(familyprop.v,"../results/tables/phylotype/conserved_taxonomy/familyprop.v.tsv", quote = FALSE, sep="\t",
            row.names = FALSE, col.names = TRUE)

write.table(genusprop.v,"../results/tables/phylotype/conserved_taxonomy/genusprop.v.tsv", quote = FALSE, sep="\t",
            row.names = FALSE, col.names = TRUE)
```

### Review dimensions of all taxa data

*Phylum*
```{r, echo=TRUE, comment=NA}
dim(phylum.a)
dim(phylum)
dim(phylum.t)
dim(phylumprop)
dim(phylum.t.v)
```

*Class*
```{r, echo=TRUE, comment=NA}
dim(class.a)
dim(class)
dim(class.t)
dim(classprop)
dim(class.t.v)
```

*Order*
```{r, echo=TRUE, comment=NA}
dim(order.a)
dim(order)
dim(order.t)
dim(orderprop)
dim(order.t.v)
```

*Family*
```{r, echo=TRUE, comment=NA}
dim(family.a)
dim(family)
dim(family.t)
dim(familyprop)
dim(family.t.v)
```

*Genus*
```{r, echo=TRUE, comment=NA}
dim(genus.a)
dim(genus)
dim(genus.t)
dim(genusprop)
dim(genus.t.v)
```

<br>

```{r, eval=FALSE, include=FALSE}
sink("./sessionInfo.txt")
message("\nSummary of session information\n\n")
sessionInfo()
sink()

sink("./LatexsessionInfo.txt")
message("\nSummary of Latex session information\n\n")
toLatex(sessionInfo())
sink()
```

